---
title: "My title"
subtitle: "My subtitle if needed"
author: Mariana Garcia Mejia
thanks: "Code and data are available at: [https://github.com/mejiama16/Toronto_Safety_Project](https://github.com/mejiama16/Toronto_Safety_Project)."
date: 08 May 2026
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# Load packages
library(tidyverse)

# Load data
df <- read_csv("analysis_data.csv")
```


# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....






# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Homicides Overtime (2011-2026)


Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.



Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-hom-overtime-total
#| fig-cap: Total homicides overtime (2011-2020)
#| echo: false

df |>
  group_by(YEAR) |>
  summarise(count = n()) |> # number of entries/homicides per year
  ggplot(aes(x = YEAR, y = count)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(2011, 2026, by = 1)) +
  labs(
    title = "Homicides per Year (2011-2020)",
    x = "Year",
    y = "Number of Homicides"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) # center
```

```{r}
#| label: fig-hom-overtime-type
#| fig-cap: Total homicides overtime by type (2011-2020)
#| echo: false
#| warning: false

df |>
  group_by(YEAR, HOMICIDE_TYPE) |>
  summarise(count = n(), .groups = "drop") |>
  ggplot(aes(x = YEAR, y = count, color = HOMICIDE_TYPE)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(2011, 2026, by = 1)) +
  labs(
    title = "Homicides per Year by Type",
    x = "Year",
    y = "Number of Homicides",
    color = "Homicide Type"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: fig-hom-type-bars
#| fig-cap: Total homicides by type (2011-2020)
#| echo: false
#| warning: false

cleaned_data |>
  group_by(HOMICIDE_TYPE) |>
  summarise(count = n(), .groups = "drop") |>
  ggplot(aes(fct_reorder(HOMICIDE_TYPE, -count), y = count, fill = HOMICIDE_TYPE)) +
  geom_col() +
  labs(
    title = "Homicides by Type (2011-2020)",
    x = "Homicide Type",
    y = "Number of Homicides",
    fill = "Homicide Type"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: fig-hom-month
#| fig-cap: Total homicides per month (2011-2020)
#| echo: false
#| warning: false

cleaned_data |>
  group_by(MONTH) |>
  summarise(count = n(), .groups = "drop") |>
  ggplot(aes(x = MONTH, y = count)) +
  geom_col() +
  labs(
    title = "Homicides per Month (2011-2020)",
    x = "Month",
    y = "Number of Homicides",
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#| label: table-top5-neighborhoods
#| tbl-cap: Five neighborhoods with the most homicides (2011-2020)
#| echo: false
#| warning: false

cleaned_data |>
  group_by(NEIGHBORHOOD) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count)) |>
  slice(1:5) |>
  knitr::kable(col.names = c("Neighbourhood", "Number of Homicides"))
```

```{r}
#| label: table-top5-neighborhoods-recent
#| tbl-cap: Five neighborhoods with the most homicides in recent years (2018-2020)
#| echo: false
#| warning: false

cleaned_data |>
  filter(YEAR == 2018 | YEAR == 2019 | YEAR == 2020) |>
  group_by(NEIGHBORHOOD) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count)) |>
  slice(1:5) |>
  knitr::kable(col.names = c("Neighbourhood", "Number of Homicides"))

```


```{r}
#| label: table-top5-divisions
#| tbl-cap: Five police divisions with the most homicides (2011-2020)
#| echo: false
#| warning: false

cleaned_data |>
  group_by(DIVISION) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count)) |>
  slice(1:5) |>
  knitr::kable(col.names = c("Division", "Number of Homicides"))
```

```{r}
#| label: table-top5-division-recent
#| tbl-cap: Five neighborhoods with the most homicides in recent years (2018-2020)
#| echo: false
#| warning: false

cleaned_data |>
  filter(YEAR == 2018 | YEAR == 2019 | YEAR == 2020) |>
  group_by(DIVISION) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count)) |>
  slice(1:5) |>
  knitr::kable(col.names = c("Division", "Number of Homicides"))

```


Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.


```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


